---
title: "Logistic Regession"
author: "F.A. Barrios"
date: "`r Sys.Date()`"
output:
  pdf_document:
  html_document:
    toc: yes
    toc_depth: 3
    number_sections: true
    toc_float:
      collapsed: false
    cod_folding: hide
    center: true
    theme: united
description: "to prepare Class2020 presentations"
---


```{r setup, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

```{r}
# All the needed libraries
library(tidyverse)
library(emmeans)
library(wesanderson)
library(rstatix)
library(HSAUR2)
library(car)
library(effects)

setwd("~/Dropbox/GitHub/Class2020")
wcgs <- read_csv("DataRegressBook/Chap2/wcgs.csv")
```
# Examples from the CAR book (Fox & Weisberg) [^1]

[^1]: All notes are taken form the "Companion to Applied Regression", 3rd Ed. Fox & Weisberg

## Review of the Structure of GLMs

The structure of a GLM is very similar to that of the linear model. In particular we have a response variable *y* and *k* predictors, and we are interested in understanding how the mean of *y* varies as the values of the predictors change.


A GLM consists of three components

1. Random component, specifying the conditional or "error" distribution of the response variable, $y$, given the predictors from an *exponential family*.  Both the binomial and Poisson distributions ae in the class of explonential families, and so problems with categorical or discrete responses can be studied with GLMs.

2. As in linear models, the $m$ predictors in a GLM are translated into a vector of $k + 1$ regresor variables, ${\bf x} = (x_0, x_1, \dots , x_k)$, possibly using contrast regressors for factors, polynomials, regression splines, transformations, and interactions. The response depends on the predictors only through a linear function of the regressors, called the *linear predictor*, $\eta({\bf x}) = \beta_0 + \beta_1 x_1 + \cdots + \beta_k x_k$.

3. The connection between the conditional mean $E[y|{\bf x}]$ of the response and the predictor $\eta({\bf x})$ in a linear model is direct,
$$E[y|{\bf x}] = \eta({\bf x}) = \beta_0 + \beta_1 x_1 + \cdots + \beta_k x_k$$
and so the mean is equal to a linear combination of the regressors. This direct relation is not appropriate for all GLM because $\eta({\bf x})$ can take any value, whereas the mean of a binary response variable must be in the interval (0,1). Therefore we introduce an invertible *link function g* that translates from the scale of the mean response to the scale of the linear predictor. $\eta({\bf x}) = E[y|{\bf x}]$ is standard in the GLM for the conditional mean of the response, therefore $g[\mu({\bf x})] = \eta({\bf x})$
Reversing this relationship produces the *inverse-link function*, $g^{-1}[\eta(\bf {x})] = \mu(\bf {x})$. The inverse of the link function is sometimes is sometimes called the *mean link funcion*


Standard link functions and their inverses table: $\mu = E[y|\bf x]$ is the expected value of the response; $\eta = \beta_0 + \beta_1 x_1 + \cdots + \beta_k x_k$ is the linear predictor.
$$\begin{array}{cccc}
  \bf Link & {\bf  \eta = g(\mu)} & {\bf \mu = g^{-1}(\eta)} & \bf Inverse Link \\
  identity & \mu & \eta & identity \\
  log  & \log(\mu) & e^\eta & exponential  \\
  inverse & \mu^{-1} & \eta^{-1} & inverse  \\
  inverse square & \mu^{-2} & \eta^{-1/2} & inverse square root  \\
  square root & \sqrt{\mu} & \eta^{2} & square \\
  logit & \log\frac{\mu}{1-\mu} & \frac{1}{1+e^{-\eta}} & logistic \\
  probit & \Phi(\mu) & \Phi^{-1}(\eta) & normal quantile \\
  comp. log-log & \log[-\log(-\mu)] & 1- \exp[-\exp(\eta)] & -
\end{array} $$


And the table for canonical or default link, response range, and conditional variance function for GLM families.
$$\begin{array}{cccc}
  \bf{Family} & {\bf Default Link} & {\bf Range of y}  & {\bf Var}(y|{\bf x}) \\
  gaussian & identity & (-\infty, +\infty) & \phi \\
  binomial & logit & \frac{0,1,\dots, N}{N} & \frac{\mu(1-\mu)}{N} \\
  poisson & log & 0,1,\dots & \mu \\
  Gamma & inverse & (0,\infty) & \phi \mu^2 \\
  Inverse.gaussian & \frac{1}{\mu^2} & (0,\infty) & \phi \mu^3
\end{array} $$

The variance distributions of an exponential family is a product of a positive *disperssion (scale)* parameter $\phi$ and a function of the mean given the linear predictor:
$$Var(y|{\bf x}) = \phi \times V[\mu({\bf x})]$$
The variances for several exponential families are listed in the table above.

The *deviance*, based on the maximized value of the log-likelihood, provides a measure of the fit of a GLM to the data, much as the residual sum of squares does for a linear model. The value of the log-likelihood evaluated at the maximum likelihood estimates the regression coefficients for fixed dispersion is
$$ \log L_0 = \sum \log p[y_i;\hat \mu({\bf x_i}), \phi] $$
An fitting to a saturated model, with one parameter for each of the *n* observations, with the mean response for each observation just the observed value 
$$ \log L_1 = \sum \log p[y_i; y_i , \phi] $$
THe resudual deviance is defined as twice the difference of the log-likelihoods,
$$D({\bf y; \hat \mu}) = 2(\log L_1 - \log L_0) $$
the larger the diviance, the less well the the model of interest matches the data.

## GLMs for Binary Responce Data

Considering data in which each case provides a *binary response*, say "success" or "failure", the cases are independent, and the probability of success $\mu({\bf x})$ is the same for all cases with the same values **x** of the regressors.

When the response is binary, we think of the mean function $\mu({\bf x})$ as the conditional probability that response is success given the values **x** of the regressors. The most common link funtction used with binary response data is the ligit link, for which
$$\log[\frac{\mu({\bf x})}{1-\mu({\bf x})}] = \eta({\bf x})$$

The left side of the equation is called the *logit* of the *log-odds*, where the *odds* are the probability of success divided by the probability of failure. Solving for 
$\mu({\bf x})$ gives the mean function,
$$\mu({\bf x}) = \frac{1}{1+\exp[-\eta{\bf x})]}$$

Plot of the comparison of the logit, probit, and complementary log-log links

```{r}
# Example from the car Book Chp 6
Probit <- binomial(link=probit)
Logit <- binomial(link=logit)
Cloglog <- binomial(link=cloglog)
range <- seq(-10,10,length=1000)
plot(range,Logit$linkinv(range),type="l", xlim=c(-5,5), lty=1,
       xlab=expression(eta(x)), ylab=expression(mu(x)))
lines(sqrt(pi^2/3)*range, Probit$linkinv(range), lty=2)
lines(range,Cloglog$linkinv(range), lty=4, lwd=2)
legend("topleft",c("logit", "probit", "cloglog"), lty=c(1,2,4),
       lwd=c(1,1,2), inset=0.02)
```

## Example: Women's Labor Force Participation

To illustrate logistic regression, we turn to a study of the U.S. Panel Study of Income Dynamics of the response variable is married women's labor force participation. The data is in Mroz (carData).
$$\begin{array}{llc}
  \bf{Variable} & {\bf Description} & {\bf Remarcs} \\
  lfp & \verb+labor force participation+ & factor: no, yes \\
  k5 & \verb+number of children ages 5 and younger+ & 0-3 \\
  k618 & \verb+number of children ages 6 to 18+ & 0-8\\
  age & \verb+wife's age in yars+ & 30-60 \\
  wc & \verb+wife's college attendance+ & factor: no, yes \\
  hc & \verb+husband's college attendance+ & factor: no, yes \\
  lwg & \verb+log of estimated wife's wage+ &  \\
  inc & \verb+family income excluding wife's income+ & 1000s
\end{array} $$

So, the estimated logistic-regression model is given by

$$log[\frac{\hat\mu(x)}{1-\hat\mu(x)}] = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \cdots + \beta_k x_k$$

If exponentiate both sides of the equation, we get
$$\frac{\hat\mu(x)}{1-\hat\mu(x)} = exp(\beta_0) \times exp(\beta_1 x_1) \times exp(\beta_2 x_2) \times \cdots \times exp(beta_k x_k)$$

where the left hand of the equation, $\frac{\hat\mu(x)}{1-\hat\mu(x)}$, gives the *fitted odds* of success, **the fitted probability of success divided by the fitted probability of failure**. Exponentiating the model removes the logarithms and changes the model in the log-odds scale to one that is multiplicative, in this log odds scale. 

```{r}
##################### From car
# carData Mroz
summary(Mroz)
# logistic model family= binomial's default link is logit
mroz.mod <- glm(lfp ~ k5 + k618 + age + wc + hc + lwg + inc, family=binomial, data=Mroz)
S(mroz.mod)
```

Exponentiating the model removes the logarithms (S function shows the exponents of the betas).  For example increasing the age of a woman by one year, holding the other predictors constant, from $odds = \exp(c_0) \times \exp(c_1) \times \exp(c_2) \times \exp(\beta_3) \times \exp(c_4) \dots$ therefore multiplies the fitted odds of her being in the workforce by $\exp(\beta_3) = \exp(-0.06287) = 0.9391$. That is, reduces the odds of working by $100(1-0.9391) \approx 6\%$.
Compared to a woman that who did not attend college, a college-educated woman with all other predictors fixed has fitted odds of working about 2.24 times higher, with a 95% confidence interval [1.43, 3.54].  The exponents of the coefficient estimates are called *risk factors* or *odds ratios*. The confidence intervals for the GLM are based on profiling the log-likelihood. The confidence intervals may not be symmetric.

## Volunteering for a Psychological Experiment

Cowles collected data on the willingness of students in an introductory psychology class to volunteer for a psychological experiment.
The data set contains several variables:
1. The personality dimension *neuroticism*, a numeric variable with integer scores on a scale from zero to 24
2. The personality dimension *extraversion*, also a numeric variable with a potential range of zero to 24.
3. The factor sex, with levels "female" and "male".
4. Tha factor volunteer, with levels "no" and "yes".

Researchers expected volunteering to depend on the sex variable and on the interaction of the personality dimensions, so included the linear-by-linear interaction between nuroticism and extraversion:

```{r}
brief(Cowles)
sum(Cowles$volunteer == "yes") # number yes

cowles.mod <- glm(volunteer ~ sex + neuroticism*extraversion,
    data=Cowles, family=binomial)
brief(cowles.mod, pvalues=TRUE)
```

## Predictor Effect Plots for Logistic Regression

The **effects** package, can draw predictor effect plots for generalized linear models, including logistic regression.
```{r}
plot(predictorEffects(mroz.mod))

plot(predictorEffects(cowles.mod, ~ neuroticism + extraversion,
    xlevels=list(neuroticism=seq(0, 24, by=8), 
        extraversion=seq(0, 24, by=8))), 
    lines=list(multiline=TRUE))
```

## Analysis of Deviance and Hypotesis Test for Logistic Regression

### Model Comparisons

```{r}
mroz.mod.2 <- update(mroz.mod, . ~ . - k5 - k618)
anova(mroz.mod.2, mroz.mod, test="Chisq")

brief(cowles.mod.0 <- update(cowles.mod, 
    . ~ . - neuroticism:extraversion))
anova(cowles.mod.0, cowles.mod, test="Chisq")

# Type II Tests 
Anova(mroz.mod)

Anova(cowles.mod)

# Other Hypothesis Tests
linearHypothesis(mroz.mod, c("k5", "k618"))

linearHypothesis(mroz.mod, "k5 = k618")
```

## Fitted and Predicted Values

The function $\verb+predict()+$ returns the estimated linear predictor values for each observation. And to get the fitted probabilities we use the argument type="response". The $\verb+fitted()+$ function can also be used.

```{r}
opts <- options(digits=5)
head(predict(mroz.mod)) # first few values

head(predict(mroz.mod, type="response"))

```

And the $\verb+predict()+$ can be used to compute predicted values for arbitrary combinations of predictor values. For example we can estimate the probability of volunteering at neuroticism and extraversion at 12.

```{r}
options(opts)

predict.data <- data.frame(sex=c("female", "male"), 
    neuroticism=rep(12, 2), extraversion=rep(12, 2))
predict.data$p.volunteer <- predict(cowles.mod, 
    newdata=predict.data, type="response")
predict.data
```


# Binomial Data

In binomial response data, the response variable $y_i$ for each case $i$ is the number of successes in a fixed number $N_i$ of independent trials, each with the same probability of success.
Binary regression is a limiting case of binomial regression with all the $N_i = 1$

$$\begin{array}{llccc}
  \bf{\verb+Perceived Closeness+} & {\bf \verb+Intensity of Preference+} & {\bf \verb+Voted+} & {\bf \verb+ Did Not Voted+} & {\bf logit}\\
  One-sided & \verb+Weak+ & 91 & 39 & 0.847 \\
  One-sided & \verb+Medium+ & 121 & 49 & 0.904 \\
  One-sided & \verb+Strong+ & 64 & 24 & 0.981 \\
  Close & \verb+Weak+ & 214 & 87 & 0.900 \\
  Close & \verb+Medium+ & 284 & 76 & 1.318 \\
  Close & \verb+Strong+ & 201 & 25 & 2.084
\end{array} $$

```{r}
Campbell <- data.frame(
    closeness = factor(rep(c("one.sided", "close"), c(3, 3)),
        levels=c("one.sided", "close")),
    preference = factor(rep(c("weak", "medium", "strong"), 2),
        levels=c("weak", "medium", "strong")),
    voted = c(91, 121, 64, 214, 284, 201),
    did.not.vote = c(39, 49, 24, 87, 76, 25)
)
Campbell
```

For binomial data, the response can be the *proportion* of successes for each observation, or the proportion of successes for failures. For example the logit for the One-sided, weak case 91 voted, and 39 did not voted, $\log(\verb+Voted/Did not Voted+) = \log(91/39)$ -> *0.847*

```{r}
campbell.mod <- glm(cbind(voted, did.not.vote) ~
    closeness*preference, family=binomial, data=Campbell)
# the estimated responses are exact and the residuals are zero
predict(campbell.mod)
residuals(campbell.mod)

plot(predictorEffects(campbell.mod, ~ preference),  
     main="", confint=list(style="bars"), lines=list(multiline=TRUE),
     xlab="Intensity of Preference", ylab="Probability of Voting",
     lattice=list(key.args = list(x =0.1, y = 0.95, corner=c(0, 1))))
```

The emmeans function can be used to test differences between the two levels of closeness fo each level of preference:

```{r}
emmeans(campbell.mod, pairwise ~ closeness | preference)$contrasts
```
