---
title: "Exampl_t_ANOVA.Rmd"
author: "F.A. Barrios"
date: "10/10/2020"
output: html_document
---

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Examples Chap03 

The examples for chapter 3 using data from the heart and estrogen/progestin study (HERS), a clinical trial of hormone therapy (HT) for prevention of recurent heart attacks and death among 2,763 post-menopausal women with existing voronary heart disease (CHD)

## Introduction
t-Test example presented in Tabel 3.1 of the t-Test of difference in average glucose by exercise for the women that are not diabetic. These eamples are to revisit som t-test R estimations 

```{r}
setwd("~/Dropbox/Fdo/ClaseStats/RegresionClass/RegresionR_code")
# To set the working directory at the user dir
library(tidyverse)
hers <- read_csv("DataRegressBook/Chap3/hersdata.csv")
# Loading the HERS database in hers varible
summary(hers)
boxplot(glucose ~ diabetes, data=hers)
# For the 
boxplot(glucose[hers$diabetes == "no"] ~ exercise[hers$diabetes == "no"], alternative="two.sided", data=hers)
t.test(glucose[hers$diabetes == "no"] ~ exercise[hers$diabetes == "no"], data=hers, alternative="two.sided", mu=0, paired=FALSE, var.equal=TRUE)
```

## Including Plots

```{r}
summary(hers$diabetes)
ggplot(data = hers, mapping = aes(x = exercise, y = glucose)) 
    + geom_boxplot() 
    + facet_grid( . ~ diabetes)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

# Example from R-bloggers
```{r}
library(multcomp)
library(tidyverse)

# Create the four groups
set.seed(10) 
df1 <- data.frame(Var="a", Value=rnorm(100,10,5))
df2 <- data.frame(Var="b", Value=rnorm(100,10,5))
df3 <- data.frame(Var="c", Value=rnorm(100,11,6))
df4 <- data.frame(Var="d", Value=rnorm(100,11,6))

# merge them in one data frame
df<-rbind(df1,df2,df3,df4)

# convert Var to a factor
df$Var<-as.factor(df$Var)
df%>%ggplot(aes(x=Value, fill=Var))+geom_density(alpha=0.5)
```
# The ANOVA
The ANOVA model and some examples
```{r}
# ANOVA
model1<-lm(Value~Var, data=df)
anova(model1)
```

# Tukey multiple comparisons
```{r}
summary(glht(model1, mcp(Var="Tukey")))
```
