---
title: "General Linear Model"
author: "F. A. Barrios<br><small>Instituto de Neurobiología UNAM<br></small>"
date: "<small>`r Sys.Date()`</small>"
output:
  html_document:
    toc: yes
    toc_depth: 3
    number_sections: true
    toc_float:
      collapsed: false
    cod_folding: hide
    theme: cerulean
  pdf_document: default
description: "to prepare Class2020 presentations"
---

---
```{r setup, echo=FALSE}
require(Hmisc)
knitrSet(lang='markdown', fig.path='png/', fig.align='left', w=6.5, h=4.5, cache=TRUE)
# If using blogdown: knitrSet(lang='blogdown')
# knitr::opts_chunk$set(echo = TRUE)
```
`r hidingTOC(buttonLabel="Outline")`

```{r}
setwd("~/Dropbox/Fdo/ClaseStats/RegressionClass/RegressionR_code")
# To set the working directory at the user dir
library(tidyverse)
library(multcomp)
library(car)
library(emmeans)
```
# General Linear Model GLM (Modelo lineal General)

## GLM
Responce variable $Y$ is a random variable that is measured and has a distribution with expected value $ E(Y|x)$ given a set of independent variables $x$.
$$Y_j (j=1, . . . , J)$$
for a set of $x_jl$ predictor variables (or independent variaables) defined as vectors for each $j$
$$ x_jl (l=1, . . . , L)$$
with $L(L<J$, a general linear model with an error function $\epsilon_j$ can be expressed:
$$Y_j = x_{j1}\beta_1 + x_{j2}\beta_2 + x_{j3}\beta_3 + . . . + x_{jL}\beta_L + \epsilon_j$$
with $\epsilon_j$ an independent variable identically distributed to the Normal with mean equal to zero.
$$\epsilon_j \approx N(0,\sigma^2)_{iid}$$
Example of simple linear regression: exercise and glucose Glucose levels above 125 mg/dL are diagnostic of diabetes, while 100-125 mg/dL signal increased risk.
Data from HERS (public data) has baseline of glucose levels among 2,032 participants in a clinical trial of hormone therapy (HT). Women with diabetes are excluded, to study if the exercise might help prevent progression to diabetes.

```{r}
hers <- read_csv(“DataRegressBook/Chap3/hersdata.csv”)
# hers data structure
hers_nodi <- filter(hers, diabetes == “no")
hers_nodi_Fit <- lm(glucose ~ exercise, data = hers_nodi)
# the linear model results can be printed using summary
summary(hers_nodi_Fit)
```

```{r}
#  Example of the HERS data for diabetic participants
hers_yesdi <- filter(hers, diabetes == "yes")
hers_yesdi <- mutate(hers_yesdi, physact = factor(physact, levels=c("much less active","somewhat less active","about as active","somewhat more active","much more active")))
#  Example of ANOVA with HERS data for diabetic participants
#
ggplot(data = hers_yesdi, mapping = aes(x = physact, y = glucose)) + geom_boxplot(na.rm = TRUE)
glucose_yesdi_act <- lm(glucose ~ physact, data = hers_yesdi)
Anova(glucose_yesdi_act, type="II")
#
S(glucose_yesdi_act)
glucose_emmeans <- emmeans(glucose_yesdi_act, "physact")
contrast(glucose_emmeans, adjust="sidak")
```

### Including Plots

```{r}
summary(hers$diabetes)
ggplot(data = hers, mapping = aes(x = exercise, y = glucose)) + geom_boxplot() + facet_grid( . ~ diabetes)
```

