---
title: "General Linear Model"
author: "F. A. Barrios<br><small>Instituto de Neurobiología UNAM<br></small>"
date: "<small>`r Sys.Date()`</small>"
output:
  html_document:
    toc: yes
    toc_depth: 3
    number_sections: true
    toc_float:
      collapsed: false
    cod_folding: hide
    theme: cerulean
  pdf_document: default
description: "to prepare Class2020 presentations"
---

---
```{r setup, echo=FALSE}
require(Hmisc)
knitrSet(lang='markdown', fig.path='png/', fig.align='left', w=6.5, h=4.5, cache=TRUE)
# If using blogdown: knitrSet(lang='blogdown')
# knitr::opts_chunk$set(echo = TRUE)
```
`r hidingTOC(buttonLabel="Outline")`

```{r}
setwd("~/Dropbox/Fdo/ClaseStats/RegressionClass/RegressionR_code")
# To set the working directory at the user dir
library(tidyverse)
library(multcomp)
library(car)
library(emmeans)

hers <- read_csv("DataRegressBook/Chap3/hersdata.csv")
```
# General Linear Model GLM (Modelo lineal General)

## GLM
Response variable $Y$ is a random variable that is measured and has a distribution with expected value $E(Y|x)$ given a set of independent variables $x$.
$$Y_j (j=1, . . . , J)$$
for a set of $x_jl$ predictor variables (or independent variables) defined as vectors for each $j$
$$ x_jl (l=1, . . . , L)$$
with $L(L<J$, a general linear model with an error function $\epsilon_j$ can be expressed:
$$Y_j = x_{j1}\beta_1 + x_{j2}\beta_2 + x_{j3}\beta_3 + . . . + x_{jL}\beta_L + \epsilon_j$$
with $\epsilon_j$ an independent variable identically distributed to the Normal with mean equal to zero.
$$\epsilon_j \approx N(0,\sigma^2)_{iid}$$
Example of simple linear regression: exercise and glucose Glucose levels above 125 mg/dL are diagnostic of diabetes, while 100-125 mg/dL signal increased risk.
Data from HERS (public data) has baseline of glucose levels among 2,032 participants in a clinical trial of Hormone Therapy (HT). Women with diabetes are excluded, to study if the exercise might help prevent progression to diabetes.

```{r}
# hers data structure
hers_nodi <- filter(hers, diabetes == "no")
hers_nodi_Fit <- lm(glucose ~ exercise, data = hers_nodi)
# the linear model results can be printed using summary
summary(hers_nodi_Fit)
```

Simple linear regression model shows coefficient estimate (Coef) for exercise shows that average baseline glucose levels were about 1.7mg/dL lower among women who exercised at least three times a week than among women who exercised less.

```{r}
#  Example of the HERS data for diabetic participants
hers_yesdi <- filter(hers, diabetes == "yes")
hers_yesdi <- mutate(hers_yesdi, physact = factor(physact, levels=c("much less active","somewhat less active","about as active","somewhat more active","much more active")))
#  Example of ANOVA with HERS data for diabetic participants
#
ggplot(data = hers_yesdi, mapping = aes(x = physact, y = glucose)) + geom_boxplot(na.rm = TRUE)
glucose_yesdi_act <- lm(glucose ~ physact, data = hers_yesdi)
Anova(glucose_yesdi_act, type="II")
#
S(glucose_yesdi_act)
glucose_emmeans <- emmeans(glucose_yesdi_act, "physact")
contrast(glucose_emmeans, adjust="sidak")
```

### More Box Plots

```{r}
summary(hers$diabetes)
ggplot(data = hers, mapping = aes(x = exercise, y = glucose)) + geom_boxplot() + facet_grid( . ~ diabetes)
```

## For a multiple linear model
There are models to regress several predictor variables to relate several random independent variables.

$$y_i = E[y_i|x_i] + \epsilon_i$$
$$Y = \beta_0 + \beta_1 x_{1} + \beta_2 x_{2} + \dots + \beta_p x_{p}$$

Multiple linear regression model coefficients, the betas, give the change in $E[Y|x]$ for an increase of one unit on the predictor $x_j$ , holding other factors in the model constant; each of the estimates is adjusted for the effects of all the other predictors. As in the simple linear model the intercept $\beta_0$ (beta zero) gives the value $E[Y|x]$ when all the predictors are equal to zero. Example of multiple linear model estimate is done with:  glucose ~ exercise + age + drinkany + BMI.

In general in R we can write:$Y = variable_1 + variable_2 + variable_3$ for a multiple linear model.
```{r}
hers_nodi_multFit <- lm(glucose ~ exercise + age + drinkany + BMI, data = hers_nodi)
# the linear model results can be printed using summary
summary(hers_nodi_multFit)
```
## Multiple linear model, with interactions
In general in R we can write:$Y = variable_1 + variable_2 + variable_1:varible_2$ for a multiple linear model.

$$Y=\beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_1 x_2$$

(The following is a very good link: http://www.sthda.com/english/articles/40-regression-analysis/ )
