---
title: "TareaANOVA.Rmd"
author: "F. A. Barrios<br><small>Instituto de Neurobiología, UNAM<br></small>"
date: "<small>`r Sys.Date()`</small>"
output:
  html_document:
    toc: yes
    toc_depth: 3
    number_sections: true
    toc_float:
      collapsed: false
    cod_folding: hide
    theme: cerulean
  pdf_document: default
description: "to prepare homeworks"
---

```{r setup, echo=FALSE}
require(Hmisc)
knitrSet(lang='markdown', fig.path='png/', fig.align='left', w=6.5, h=4.5, cache=TRUE)
# If using blogdown: knitrSet(lang='blogdown')
# knitr::opts_chunk$set(echo = TRUE)
```
`r hidingTOC(buttonLabel="Outline")`

# Chapter 8 Daniel

Solution of some examples and exercises from Daniel's book, the ANOVA chapter.

## Including Code and comments

As part of the Homework we include R code in the document, and the R code will be included too.

# 1. Example 8.2.1

```{r}
setwd("~/Dropbox/Fdo/ClaseStats/RegressionClass/RegressionR_code")
# Example Chap 8, One-Way ANOVA
Exa8_2=read.csv(file="DataOther/EXA_C08_S02_01.csv", header=TRUE)
summary(Exa8_2)
boxplot(Exa8_2)
Exa8_2.dat = stack(Exa8_2)
names(Exa8_2.dat)
Exa8_2.aov=aov(values~ind, data=Exa8_2.dat)
summary(Exa8_2.aov)
# # # # # # # # # # # # # # # # # 
plot(values~ind, data=Exa8_2.dat)
TukeyHSD(Exa8_2.aov, ordered=TRUE)
plot(TukeyHSD(Exa8_2.aov), las=2)
abline(v=0, col=2)
```

# Including Plots

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
You can also plot on different quadrants, for example:

```{r, echo=FALSE}
# Advanced graph.
layout(matrix(c(1, 2, 3, 4), nrow=2, ncol=2, byrow=TRUE))
plot(Exa8_2.aov)
```

# 2. Example 8.3.1

The section 2 of the ANOVA chapter in Daniel's book is for a two-way ANOVA. The data for this example is not loadable from a data file provided by the site of the book. Nevertheless, we will make a table copying the data directly, which is not large, form the text.

```{r}
# Build a table, index by index since
# Two-Way ANOVA EXAMPLE 8.3.1 has NO-data file
# Therefore we make the example table.
#
Age=factor(rep(c(1,2,3,4,5), 3))
Age
Method=factor(rep(1:3,c(5,5,5)))
Method
levels(Method)=letters[1:3]
Method
t = c(7,8,9,10,11,9,9,9,9,12,10,10,12,12,14)
#
plot(t ~ Age + Method)
# now the ANOVA model
t_aov = aov(t ~ Method + Age)
summary(t_aov)
layout(matrix(c(1,2), nrow=2, ncol=1, byrow=TRUE))
TukeyHSD(t_aov)
plot(TukeyHSD(t_aov))
abline(v=0, col=2)
```
# Pltotting
Looking at man of rep, levels Advanced graph.

```{r}
layout(matrix(c(1, 2, 3, 4), nrow=2, ncol=2, byrow=TRUE))
plot(t_aov)

```




