---
title: "Splines for Estimating Shape of Regression Functions"
author: "F.A. Barrios"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
---


```{r setup, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

```{r}
library(tidyverse)
library(Hmisc)
library(rms)
library(splines)
library(car)
library(MASS)
library(lspline)
#
setwd("~/Dropbox/GitHub/Class2020")
# loading the HERS data set from Vittinghoff's data
hers <- read_csv("DataRegressBook/Chap3/hersdata.csv")
```

## Linear Splines

The simplest spline function is the linear spline function, a piecewise linear function. The assumption is that the $x$ axis is divided into intervals wit hend points at $a, b,$ and $c$, called knots. The linear spine function is defined by
$$ f(X) = \beta_0 + \beta_1 X + \beta_2 (X-a)_+ + \beta_3 (X-b)_+ + \beta_4 (X-c)_+ $$
where
$$ (u)_+  =  u, u>0,$$
$$ = 0, u \leq 0 $$
The general linear model can be written assuming only piecewise linearity in X

$$ X \beta = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3 + \beta_4 X_4 $$
and
$$ X_1 = X , X_2 = (X-a)_+ , X_3 = (X-b)_+ , X_4 =(X-c)_+ $$

## Cubic Spline Functions

Cubic polinomials have been found tohave nice properties with good ability to fit sharply curving shapes.  Cubic splines can be made to be smooth at the joint points(knots), by forcing the first and second derivatives of the function to agree at the knots:
$$ f(X) = \beta_0 + \beta_1 X + \beta_2 X^2 + \beta_3 X^3 + \beta_4 (X-a)^{3}_+ + \beta_5 (X-b)^{3}_+ + \beta_6 (X-c)^{3}_+ $$

## Restricted Cubic Splines

Cubic splines behave poorly at the tails, that is before the first knot and after the last knot. Natural splines $ns$ or restricted cubic spline function constrain the function to be linear at the tails

$$f(X) = \beta+0 + \beta_1 X_1 + \beta_2 X_2 + \cdots + \beta_{k-1} X_{k-1}$$


where $X_1 = X$ and for $j= 1, \dots ,k-2$,


$$ X_{j+1} = (X - t_j)^{3}_+ - (X - t_{k-1})^{3}_+(t_k - t_j)/(t_k - t_{k-1}) + (X - t_k)^{3}_+(t_{k-1} - t_j)/(t_k - t_{k-1}) $$
in the Hmisc and rms functions the terme in the last equation are scaled divided by $\tau = (t_k - t_1)^2$

For the HERS data ploting the LDL and HDL variables vs. BMI we get

```{r}
# to prepare the HERS data
hers <- mutate(hers, statins = factor(statins))
hers <- mutate(hers, nonwhite = factor(nonwhite))
hers <- mutate(hers, smoking = factor(smoking))
hers <- mutate(hers, drinkany = factor(drinkany))
hers <- mutate(hers, BMIc = BMI - mean(BMI, na.rm=TRUE))
hers_no_statins <- mutate(hers, statins = "no", na.rm=TRUE)
hers_statins <- mutate(hers, statins = "yes", na.rm=TRUE)
#
#
ggplot(data = hers, mapping=aes(x=BMI, y=LDL)) +
  labs(x="BodyMassIndex", y= "LDL", title = "Smooth") + 
  geom_point(shape=20, color=2) + 
  geom_smooth()

# now with splines
ggplot(data = hers, mapping=aes(x=BMI, y=LDL)) +
  labs(x="BodyMassIndex", y= "LDL", title = "Linear model") + 
  geom_point(shape=20, color=2) + 
  geom_smooth(method="lm", formula=y~x)
#
# with the splines R package functions ns Natural Splines
ggplot(data = hers, mapping=aes(x=BMI, y=HDL)) + 
  labs(x="Body Mass Index kg/m^2", y= "HDL Cholesteron mg/dL", title = "Natural Splines") + 
  geom_point(shape=20, color=2) + 
  geom_smooth(method="lm", formula=y~ns(x))
#
#
ggplot(data = hers, mapping=aes(x=BMI, y=LDL)) + labs(x="Body Mass Index kg/m^2", y= "LDL Cholesteron mg/dL", title = "Natural Splines") + 
  geom_point(shape=20, color=2) + 
  geom_smooth(method="lm", formula=y~ns(x))
#
ggplot(data = hers, mapping=aes(x=BMI, y=LDL)) + labs(x="Body Mass Index kg/m^2", y= "LDL Cholesteron mg/dL", title = "Linear Splines") + 
  geom_point(shape=20, color=4) + 
  geom_smooth(method="lm", formula=y~lspline(x, quantile(x, c(0, .05, .275, .5, .775, .95, 1), include.lowest = TRUE)))

#
ggplot(data = hers, mapping=aes(x=BMI, y=LDL)) + labs(x="Body Mass Index kg/m^2", y= "LDL Cholesteron mg/dL", title = "Linear Splines") + 
  geom_point(shape=20, color=4) + 
  geom_smooth(method="lm", formula=y~rcs(x, quantile(x, c(0, .05, .275, .5, .775, .95, 1), include.lowest = TRUE)))
              
# with the B-spline function knots = 3 form the R slines
ggplot(data = hers, mapping=aes(x=BMI, y=HDL)) + labs(x="Body Mass Index kg/m^2", y= "HDL Cholesteron mg/dL", title = "Restricted Cubic Splines") + 
  geom_point(shape=20, color=2) + 
  geom_smooth(method="lm", formula=y~bs(x, df = 3))

ggplot(data = hers_statins, mapping=aes(x=BMI, y=HDL)) + labs(x="Body Mass Index kg/m^2", y= "HDL Cholesteron mg/dL", title = "Restricted Cubic Splines 7 knots") +
  geom_point(shape=20, color=2) + 
  geom_smooth(method="lm", formula=y~rcspline.eval(x, quantile(x, c(0, .05, .275, .5, .775, .95, 1), include.lowest = TRUE)))
```

