---
title: "emmeans Notes"
author: "F.A. Barrios"
date: "`r Sys.Date()`"
output:
  rmdformats::material:
    highlight: kate
  pdf_document:
description: "to prepare Class2020 presentations"
---


```{r setup, include=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

```{r}
library(tidyverse)
library(car)
library(emmeans)
library(effects)
```

# Notes from emmeans [^1]

[^1]: Notes taken form Chap 4 of Fox & Weisberg

In a linear model with one factor, a One-Way Analysis of Variance. In the example of Baumann that conducted an experiment with 66 children, assigning at random to one of three experimental groups. The groups represent different different methods of teaching reading: a standard `Basal` and two new methods called `DTRA` and `Strat`. With two pre-tests and three post-tests of reading and comprehension.

```{r}
summary(Baumann[, c(1, 6)]) 

xtabs(~ group, data=Baumann)
```

xtabs in a one sided formula "counts" the number of cases.

```{r}
Tapply(post.test.3 ~ group, mean, data=Baumann)
Tapply(post.test.3 ~ group, sd, data=Baumann)

plot(post.test.3 ~ group, data=Baumann, xlab="Group",
    ylab="Reading Score")
```

The means and boxplots suggest that there may be systematic differences in levels among the groups. The one-way ANOVA model can determine if the difference is significative

```{r}
S(baum.mod.1 <- lm(post.test.3 ~ group, data=Baumann))

```



The pairwise comparisons of the means for the three groups can be obtained using the `emmeans` function. emmeans comes form "estimated marginal means", the *emmeans* package enables users to easily obtain least-squares means for many linear, generalized linear, and mixed models as well as compute contrasts or linear functions of least-squares means, and comparisons of slopes.

```{r}
(emms.est1.baum.mod.1 <- emmeans(baum.mod.1, pairwise ~ group))

(emms.est2.baum.mod.1 <- emmeans(baum.mod.1, trt.vs.ctrl ~ group))

plot(emms.est1.baum.mod.1)

plot(emms.est2.baum.mod.1)

```


